---
title: "Final Individual Summary"
author: "Eveline Lindqvist Nilsson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  html:
    theme: readable
    css: styles.css
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
  echo: true
---


```{r}
library(readxl)    # for Excel files
library(dplyr)     # for data wrangling
library(ggplot2)   # for plotting

```


###  Introduction

This report studies the relationship between climate and happiness. The main question is whether people who live in warmer countries are happier. Many people believe that warm weather and sunshine improve well being. However, happiness depends on many factors, such as income, health, social support, and government policies. Because of this, climate alone may not fully explain why some countries are happier than others.


In this analysis, I compare national happiness scores with average annual temperature across countries. Countries are grouped into three climate categories. Cold countries have an average temperature below about 10 degrees Celsius. Moderate countries have temperatures between 10 and 20 degrees Celsius. Warm countries have temperatures above 20 to 25 degrees Celsius. This grouping helps to identify general patterns and also shows important exceptions.



The goal of this report is to understand whether warmer countries are happier on average and whether very cold or very hot climates are linked to lower happiness. This individual report supports the group project by focusing on environmental factors instead of economic or political factors. Together, these analyses help explain what makes countries happier.


###  Data Acquisition

The happiness data comes from the World Happiness Report and covers the years from 2019 to 2025. This dataset includes national life satisfaction scores based on surveys where people rate their overall life satisfaction on a scale from 0 to 10. The World Happiness Report is widely used and is considered a reliable measure of happiness across countries.


The climate data comes from the World Bank Climate Data and includes country level temperature information. To measure climate, I use average annual temperature. This reflects long term climate conditions rather than short term weather changes.


Both datasets were downloaded as CSV files and saved in the project folder. They were loaded using relative file paths so that the analysis can be reproduced on another computer. The two datasets were merged by country name to create one dataset that includes both happiness scores and temperature values.



### Get dataset for happiness ranking 2019-2025

World Happiness Report
- [World Happiness Report](https://www.worldhappiness.report/)

World Bank Climate Data Maximum Temperature
- [World Bank Max Temperature Data](https://data360.worldbank.org/en/indicator/WB_CCKP_TASMAX?view=datatable)

World Bank Climate Data Minimum Temperature
- [World Bank Min Temperature Data](https)

The maximum and minimum temperature datasets were combined to estimate the average annual temperature for each country.

###  Two datasets were used for this analysis:

**Happiness data** from the World Happiness Report (2019–2025), which includes national life satisfaction scores and related factors.

**Average annual temperature** by country from a publicly available climate dataset.

```{r}


```


Both datasets were stored as CSV files in the project folder and loaded using relative paths to ensure reproducibility. The two datasets were merged by country, creating a unified dataset containing both temperature and happiness scores for analysis.

###  Data Cleaning and Preparation

Before starting the analysis, the data needed to be cleaned and prepared. First, the happiness data was limited to the years 2019 to 2025 so it matched the available climate data. Countries with missing happiness scores or missing temperature values were removed.


Country names were checked and adjusted so they matched across both datasets. After merging the datasets by country, I reviewed summary statistics and basic plots to check for unusual values or data errors.


To make the analysis easier to understand, countries were grouped into three climate categories. These categories are cold, moderate, and warm, based on their average annual temperature. This allows for clearer comparison of happiness levels across different climate types.

















###  Data Sources
For this analysis, we used several trusted global datasets, including the World Happiness Report, the World Bank, OECD, and Our World in Data. These sources give information about national happiness, economic conditions, social support, and environmental factors. The happiness scores are mostly self-reported, which means people rate how satisfied they feel with their lives. This makes the data subjective, but it still gives a good overall picture of well-being across countries. The temperature data comes from historical climate records, giving average annual temperatures for each country. The combination of these datasets allows us to compare environmental conditions with how happy people are. One strength of this approach is that we can look at many countries over multiple years, which gives a broader perspective. However, some data is missing for certain countries or years, and happiness is influenced by many other factors besides climate. Therefore, our results should be interpreted carefully.





**Research Question**

The main question in this report is whether people living in warmer countries are happier. Many people think that sunshine and warmth make life better, but happiness is influenced by many social, economic, and environmental factors. To investigate this, we compared countries in three temperature categories: cold, moderate, and warm. Cold countries are those with average annual temperatures below about 10 degrees Celsius. Moderate countries have temperatures between 10 and 20 degrees, and warm countries have temperatures between 25 and 30 degrees. We used several plots to examine the relationship between temperature and happiness and to see if climate alone can explain differences in happiness across countries.







**Step 1: Temperature and Happiness Across All Countries**

```{r}
```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(here)

# Load data using relative path
happiness <- read_excel(here("data", "WHR24_Data_Figure_2.1.xlsx"))
temperature <- read_excel(here("data", "temperature.xlsx"))

# Keep only 2024 column
temp_2024 <- temperature %>%
  select(code, `2024-07`) %>%
  rename(CountryCode = code, AvgTemp = `2024-07`)

# Create country lookup
country_lookup <- data.frame(
  CountryCode = c("FIN","DNK","ISL","SWE","ISR","NLD","NOR","LUX","CHE","AUS","NZL","CRI","KWT","AUT","CAN","BEL","IRL","CZE","LTU","GBR","SVN","ARE","USA","DEU","MEX","URY","FRA","SAU","KOS","SGP","TWN","ROU","SLV","EST","POL","ESP","SRB","CHL","PAN","MLT","ITA","GTM","NIC","BRA","SVK","LVA",



```


The first analysis looks at all countries together, plotting average annual temperature against happiness scores. From this plot, we notice an interesting pattern: many of the happiest countries are cold countries, often below 10 degrees Celsius. These include Nordic countries like Norway and Iceland. On the other hand, very warm countries, with average temperatures around 25 to 30 degrees, tend to have lower happiness scores. Countries with moderate temperatures show a wide range of happiness values, and there is no clear trend in one direction. This suggests that temperature alone does not explain happiness. The plot also shows that extremely warm or extremely cold countries are fewer in number, which makes their effect on the overall pattern smaller.








**Step 2: Distribution of Temperature and Happiness**

The next step was to look at the distributions of temperature and happiness separately. One plot shows how many countries fall into each temperature range. Most countries are in the moderate climate range, around 10 to 20 degrees. Another plot shows the distribution of happiness scores. Most countries have happiness scores between 4 and 7, with very few countries above 7 or below 4. These distributions show why the overall relationship between temperature and happiness is weak. There are extremely happy and unhappy countries at different temperature levels, which means that other factors, such as income, social support, and governance, play a stronger role than temperature.







**Step 3: Comparing the Warmest Countries**

To explore further, we created a plot for the five warmest countries. These countries have average temperatures close to 28 degrees. The plot shows that these countries generally have lower happiness scores. For example, Burkina Faso and Mali have temperatures around 28 degrees, but their happiness scores are around 4 to 5. However, it is important to remember that these countries also face challenges like lower economic development, weaker healthcare systems, and limited social support. This means that low happiness is likely caused by many social and economic factors, not just temperature. The plot helps us see how temperature alone is not a strong predictor of happiness.






**Step 4: Comparing the Coldest Countries**

Next, we created a plot for the five coldest countries, with average temperatures between minus 5 and 1.5 degrees Celsius. Many of these countries, like Iceland and Norway, have high happiness scores. Even very cold countries like Russia and Canada show medium happiness scores, around 5 to 6. The plot shows that cold temperatures do not prevent high happiness. It also highlights that cold countries often have strong economies, good healthcare, and strong social support, which likely contribute to their happiness. This comparison with warm countries strengthens the conclusion that climate alone does not determine happiness.





**Step 5: Mid-Temperature Countries**

We also looked at some mid-temperature countries, like Spain, Japan, Argentina, and Poland. Their temperatures are between 8.5 and 14.5 degrees Celsius, and their happiness scores range from 5.9 to 6.4. This shows that countries with moderate climates can have both high and low happiness scores. Again, this supports the idea that temperature is not the main factor influencing happiness. Other economic, social, and cultural factors are likely more important.




**Step 6: Changes Over Time for Warm Countries**

Finally, we looked at happiness over time for the warmest countries. Using multiple years of data, we can see that happiness scores in warm countries like Burkina Faso and Senegal do not change much with small changes in temperature. This confirms that temperature itself is not strongly connected to happiness over time. Long-term trends in happiness are more likely linked to changes in economic conditions, governance, or social support.




**Conclusion**

From all the plots and analyses, we can conclude that temperature does not strongly predict happiness. Warm countries are not automatically happier, and cold countries are not less happy. Countries with moderate temperatures show the full range of happiness scores. The results suggest that other factors, such as income, social support, freedom, healthcare, and governance, are much more important than climate in determining national happiness. Temperature may have some influence, but it is only one small factor among many.